\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2019-05-09 paper/response_to_referee.txt

    The paper presents a method to estimate spectrophotometric parallaxes
    based only on input data, without adoption of physically motivated
    models and priors. The paper is written in an way that is strongly
    focused on the method itself, but leaves several complementary aspects
    that are certainly important to assess the reliability of the inferred
    parallaxes and, most importantly, distances. I recommend its
    publication only after such major issues are addressed.

    Major issues

    1 - The data-driven method described in the paper details out how
    spectrophotometric parallaxes are inferred. But the paper does not
    give the reader a clue about how spectrophotometric distances are
    estimated. The authors make it clear in Section 2 that they never
    invert *observed* parallaxes to estimate distances in their
    method. But are the spectrophotometric distances they infer (shown in
    Figures 5 and 6) just the inverse of the *inferred* parallaxes? If so,
    this should be made explicit in the text, along with some discussion
    about the problems (or lack of) that may arise from this
    inversion. Even though the spectrophotometric parallaxes themselves
    are always positive, a reader would like to be convinced that the
    distances shown in Figures 5 and 6 do not suffer from biasses caused
    by the asymmetric nature of the distances distribution that results
    from a symmetric error distribution in parallaxes (as stated by the
    likelihood assumption of the method). How do these biasses simply go
    away when the method's w(sp) +/- sigma(sp) are used? In particular, a
    negative parallax for many stars may be consistent with their w(sp)
    +/- sigma(sp).

These issues are subtle! We are assuming that the GAIA PARALLAXES have
noise that is symmetric in parallax space. In our data-driven model,
we use a linear combination of features to predict or estimate a LOG
PARALLAX, so if anything we are claiming that our estimates (not the
data, but our estimates based on the data) are symmetric in log
parallax, not parallax. But we aren't even really claiming that. The
issues are subtle. However, this does deserve some discussion and we
have added that to the captions of Figure 5 and 6, and in the last
paragraph of the discussion, for good emphasis [HOGG DO THIS AT
"PLACEHOLDER A"].  The short answer is that we just exponentiated the
negative of our log-parallax estimates to make Figures 5 and 6.

There is an additional question of whether we are somehow biasing
things when we presume that the parallax must be positive? Our
position is that we are biasing things by presuming that the LOG
PARALLAX can be predicted as a linear combination of features. Of
course any other assumption would also introduce biases! There is no
way to fit that would be unbiased here, because by construction, our
model is not physically accurate. We added words about this into the
linearity assumption in the assumption list in Section 2.

@@

We will do a final URL for the repo on acceptance; we can't do it
before, because git hash! Happy to work to ensure reproducibility!

    - Do the authors wish to see the full Table 1 hosted with the Journal?
      We'd gladly process the final table into the final article, which
      would speed ingest of that data into CDS/Vizier. Regardless the
      Zenodo URL in footnote # should be converted back to its DOI form.

Yes we would like this. Let's discuss at acceptance.

In addition to the above we made a few other small changes:

- We added a reference to the Leung paper that followed ours in the
relevant part of the discussion (where we discuss machine-learning
alternatives).

### 2019-05-09 paper/sp.tex
a few things done

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
\begin{description}
\item[features]\footnote{Here and everywhere in this \documentname, we
  use the word ``feature'' to refer to an input or predictor or
  independent variable for the regression.
  That is, our parallax predictions will be linear combinations of features.
  This is not common terminology in astrophysics, but it is standard in the machine-learning literature.}
Perhaps our most fundamental assumption is that the parallax
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ (\citealt{apogee}) spectrum,
plus the $G$, $\BP$, and $\RP$ photometry from \gaia\ (\citealt{gaia}),
plus the $J$, $H$, and $K_s$ photometry from \zmass\ (\citealt{zmass}),
plus the $W_1$ and $W_2$ photometry from \wise\ (\citealt{wise}).
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

\item[linearity] Perhaps the most restrictive assumption of this work is that
the logarithmic parallax (or, equivalently, the distance modulus) 
can be predicted as a \emph{completely linear function} of
the chosen features. We are only assuming this linearity in a small range of stellar
surface gravity, but this assumption is strong, and limits strongly the
flexibility or capacity of the model.
We make it to ensure that our method is easy to optimize, and the results are easy
to interpret.
It is also the case that linear models extrapolate better than higher-order or more
flexible models; this is not extremely important to the present work, but it does
protect our predictions at the edges (in terms of temperature, surface gravity,
and chemical abundances) of our sample.

\item[likelihood] We assume that the \gaia\ parallaxes are generated by a particular
stochastic process, in which the difference between the \gaia-Catalog parallax (plus
a small offset, which we learn below) and the true parallax is effectively drawn from a
Gaussian with a width set by the \gaia-Catalog uncertainty on the parallax.
This is the standard assumption in all properly probabilistic \gaia-Mission inferences
to date, but it subsumes a number of related assumptions, like that the \gaia\ noise
model is correct, that the stars are only covariant with one another at negligible levels, and that
there are no significant outliers or large-scale systematics.
\end{description}

@@

It is interesting and valuable to compare the spectrophotometric parallax precision to
the astrometric parallax precision.
\figurename~\ref{fig:precision} compares uncertainties between the spectrophotometric
outputs and the astrometric inputs
(and recall that information is proportional to the inverse square of the uncertainty).
The brief summary is that the spectrophotometric parallaxes are more precise (and even
far more precise) than the astrometric parallaxes for stars further than a few kpc from
the Sun.
This is not unexpected; \gaia\ parallax uncertainties are in the 0.1-mas range, whereas
our uncertainties are in the ten-percent range.
\begin{figure}
\centering
~\hfill\includegraphics[width=0.8\textwidth]{./precision.pdf}\hfill~
\caption{Comparison of precisions between the spectrophotometric parallaxes and
the photometric parallaxes, as a function of magnitude and heliocentric distance.
Note that the spectrophotometric estimates are more precise for most stars.
The information content of a parallax estimate (or any measurement)
is proportional to the inverse-square of the uncertainty.\label{fig:precision}}
\end{figure}

@@

As we said above,
we required that the parallax prediction be constructed from a linear combination
of the input data or feature vector components.
This may seem like an absurd simplification; it begs the questions:
Why did it work? And could we have done better?
The answer to the first is that we have taken such a small part of stellar parameter
space with our surface-gravity cut, that the linearized model of stellar
spectrophotometry is not a bad approximation.
Also, because our linear model included photometry in magnitudes (that is, logarithmic)
and was exponentiated to make the parallax prediction, we knew in advance that much
of the needed flexibility (for dust attenuation, for example) would be close to
linear.
That is, it was a combination of limited model scope and some cleverness in the
feature engineering and model structure.

The answer to the second question is that almost certainly we could have done better!
Since models like Gaussian processes and deep networks subsume the linear model
(at least approximately), they could have delivered better---or at least non-worse---results.
The issues with going to a much more complex model are manifold, however:
We would have had many more decisions to make and many more hyper-parameters
to set.
We would have used some (or maybe much) of the information in the data to learn
the simple fact that the problem is close to linear; that is, if you give a model
a lot of freedom or capacity, it has to use a lot of the data to learn what part
of that capacity it really needs.
A more complex model might have had edge issues: Very flexible models don't extrapolate
outside the convex hull of their training data, and our training data was not
balanced in terms of parallax and dust attenuation.
In the  linear model it is also---unlike in more flexible models---trivial to propagate
uncertainties from the feature space to the prediction.
We used that, above, to propagate uncertainty in \equationname~(\ref{eq:unc}).
And finally---in principle---linear models are better for visualization and interpretation;
the linear model contains essentially only first derivatives in the data space, which are
relatively easy to understand.

@@

This \documentname\ and project delivers greatly improved distance estimates for
luminous red-giant stars, more luminous than the red clump.
The purpose of all this is to make global maps of the Milky Way.
Just as a demonstration, we show kinematics and element abundances for stars as a function
of position in the Milky Way disk in \figurename~\ref{fig:disk}.
In this visualization, it is possible to see the rotation of the disk and the
kinematic center of the Galaxy; that might even provide a distance estimate to the Galactic
Center.
It also shows extremely strong chemical gradients.
\begin{figure}
\includegraphics[width=\textwidth]{map.pdf}
\caption{A map of the kinematics and element abundances in the Milky Way disk.
  Each arrow represents the three-space Galactocentric velocity of each star
  in the sample, projected onto the Galactic disk plane.
  To make these velocities, raw \gaia\ proper motions and \apogee\ radial velocities
  were used (in a standard Galactocentric transformation), so they are
  not in any way noise-deconvolved.
  Each arrow has a color corresponding to the \apogee\ reported metallicity.
  This map shows the rotation of the Galaxy out to large radius, and also
  indicates the kinematic center of the disk rotation. The stars shown in this
  map have distance uncertainties on the order of ten percent, so at significant
  heliocentric distances, the interpretation of
  the quantitative information in this plot requires care.\label{fig:disk}}
\end{figure}
In a companion paper (\citealt{eilers}), we are using this information to estimate
the (asymmetric-drift-corrected) circular-velocity curve for the Milky Way.

Finally, one comment on the use of these spectrophotometric parallaxes.
They are noisy.
Precise use of these involves building a likelihood function and performing inference,
just as the precise use of the \gaia\ data presented here required the same.
We recommend making analogous inferences when using these outputs.

@@

>>>>

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
\begin{description}
\item[features]\footnote{Here and everywhere in this \documentname, we
  use the word ``feature'' to refer to an input or predictor or
  independent variable for the regression.
  That is, our parallax predictions will be linear combinations of features.
  This is not common terminology in astrophysics, but it is standard in the machine-learning literature.}
Perhaps our most fundamental assumption is that the parallax or distance
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ (\citealt{apogee}) spectrum,
plus the $G$, $\BP$, and $\RP$ photometry from \gaia\ (\citealt{gaia}),
plus the $J$, $H$, and $K_s$ photometry from \zmass\ (\citealt{zmass}),
plus the $W_1$ and $W_2$ photometry from \wise\ (\citealt{wise}).
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

\item[linearity] Perhaps the most restrictive assumption of this work is that
the logarithmic parallax (or, equivalently, the logarithmic distance, or,
equivalently, the distance modulus) 
can be predicted as a \emph{completely linear function} of
the chosen features. We are only assuming this linearity in a small range of stellar
surface gravity, but this assumption is strong, and limits strongly the
flexibility or capacity of the model.
We make it to ensure that our method is easy to optimize, and the results are easy
to interpret.
It is also the case that linear models extrapolate better than higher-order or more
flexible models; this is not extremely important to the present work, but it does
protect our predictions at the edges (in terms of temperature, surface gravity,
and chemical abundances) of our sample.

That said, the choice of this model (log parallax or log distance is a
linear combination of input features) will certainly introduce
particular and hard-to-learn biases into our results.
Indeed, \emph{any} choice of model that is not strictly physically
correct will introduce particular biases.
We are effectively assuming that we can produce useful distance
estimates nonetheless.

\item[likelihood] We assume that the \gaia\ parallaxes are generated by a particular
stochastic process, in which the difference between the \gaia-Catalog parallax (plus
a small offset, which we learn below) and the true parallax is effectively drawn from a
Gaussian with a width set by the \gaia-Catalog uncertainty on the parallax.
This is the standard assumption in all properly probabilistic
\gaia-Mission inferences to date (as discussed in \citealt{gaialf}),
but it subsumes a number of related assumptions, like that the
\gaia\ noise model is correct, that the stars are only covariant with
one another at negligible levels, and that there are no significant
outliers or large-scale systematics.
\end{description}

@@

It is interesting and valuable to compare the spectrophotometric parallax precision to
the astrometric parallax precision.
\figurename~\ref{fig:precision} compares uncertainties between the spectrophotometric
outputs and the astrometric inputs
(and recall that information is proportional to the inverse square of the uncertainty).
The brief summary is that the spectrophotometric parallaxes are more precise (and even
far more precise) than the astrometric parallaxes for stars further than a few kpc from
the Sun.
This is not unexpected; \gaia\ parallax uncertainties are in the 0.1-mas range, whereas
our uncertainties are in the ten-percent range.
\begin{figure}
\centering
~\hfill\includegraphics[width=0.8\textwidth]{./precision.pdf}\hfill~
\caption{Comparison of precisions between the spectrophotometric parallaxes and
the photometric parallaxes, as a function of magnitude and heliocentric distance.
For distances here are shown simply the inverses of the spectrophotometric parallaxes
$\sparallax_n$.
Note that the spectrophotometric estimates are more precise for most stars.
The information content of a parallax estimate (or any measurement)
is proportional to the inverse-square of the uncertainty.\label{fig:precision}}
\end{figure}

@@

As we said above,
we required that the log-parallax (or, equivalently, log-distance)
prediction be constructed from a linear combination
of the input data or feature vector components.
This may seem like an absurd simplification; it begs the questions:
Why did it work? And could we have done better?
The answer to the first is that we have taken such a small part of stellar parameter
space with our surface-gravity cut, that the linearized model of stellar
spectrophotometry is not a bad approximation.
Also, because our linear model included photometry in magnitudes (that is, logarithmic)
and was exponentiated to make the parallax prediction, we knew in advance that much
of the needed flexibility (for dust attenuation, for example) would be close to
linear.
That is, it was a combination of limited model scope and some cleverness in the
feature engineering and model structure.

The answer to the second question is that almost certainly we could have done better!
Since models like Gaussian processes and deep networks subsume the linear model
(at least approximately), they could have delivered better---or at least non-worse---results.
And indeed, after the the appearance of this manuscript, such an investigation
has appeared (\citealt{leung}) and it does produce very good results.
The issues with going to a much more complex model are manifold, however:
We would have had many more decisions to make and many more hyper-parameters
to set.
We would have used some (or maybe much) of the information in the data to learn
the simple fact that the problem is close to linear; that is, if you give a model
a lot of freedom or capacity, it has to use a lot of the data to learn what part
of that capacity it really needs.
A more complex model might also have edge issues: Very flexible models don't extrapolate
outside the convex hull of their training data, and our training data was not
balanced in terms of parallax and dust attenuation.
In the  linear model it is also---unlike in more flexible models---trivial to propagate
uncertainties from the feature space to the prediction.
We used that, above, to propagate uncertainty in \equationname~(\ref{eq:unc}).
And finally---in principle---linear models are better for visualization and interpretation;
the linear model contains essentially only first derivatives in the data space, which are
relatively easy to understand.

@@

This \documentname\ and project delivers greatly improved distance estimates for
luminous red-giant stars, more luminous than the red clump.
The purpose of all this is to make global maps of the Milky Way.
Just as a demonstration, we show kinematics and element abundances for stars as a function
of position in the Milky Way disk in \figurename~\ref{fig:disk}.
In this visualization, it is possible to see the rotation of the disk and the
kinematic center of the Galaxy; that might even provide a distance estimate to the Galactic
Center.
It also shows extremely strong chemical gradients.
\begin{figure}
\includegraphics[width=\textwidth]{map.pdf}
\caption{A map of the kinematics and element abundances in the Milky Way disk.
  Each arrow represents the three-space Galactocentric velocity of each star
  in the sample, projected onto the Galactic disk plane.
  To make these velocities, raw \gaia\ proper motions and \apogee\ radial velocities
  and (for distance) inverses of spectrophotometric parallaxes were used
  (in a standard Galactocentric transformation), so the map shown here is
  not in any way noise-deconvolved.
  Each arrow has a color corresponding to the \apogee\ reported metallicity.
  This map shows the rotation of the Galaxy out to large radius, and also
  indicates the kinematic center of the disk rotation. The stars shown in this
  map have distance uncertainties on the order of ten percent, so at significant
  heliocentric distances, the interpretation of
  the quantitative information in this plot requires care.\label{fig:disk}}
\end{figure}
In a companion paper (\citealt{eilers}), we are using this information to estimate
the (asymmetric-drift-corrected) circular-velocity curve for the Milky Way.

Finally, one comment on the use of these spectrophotometric parallaxes.
They are noisy.
[HOGG PLACEHOLDER A]
Precise use of these involves building a likelihood function and performing inference,
just as the precise use of the \gaia\ data presented here required the same.
We recommend making analogous inferences when using these outputs.

@@

\bibitem[Leung \& Bovy(2019)]{leung} Leung, H.~W., Bovy, J., 2019, arXiv:1902.08634
